# Introduction {#intro}

```{r, include = FALSE}
library(readr)
library(quantmod)
library(zoo)
library(tidyquant)
library(tidyverse)
library(lubridate)
library(reshape2)
library(dplyr)
library(wbstats)
library(ggplot2)
library(usmap)
library(gtrendsR)
library(tmap)
library(spData)
library(pwt9)
library(googleVis)


```
	
```{r, echo = FALSE}
unems <- wb(country = c("BRA","IND","MEX","USA"),
               indicator = c("SL.UEM.TOTL.ZS"), mrv=30, gapfill=TRUE, POSIXct = TRUE)

ggplot(unems, aes(x = date_ct, y = value, color = iso2c)) + 
  geom_line(size = 1) +
  labs(title = "U.S., Brazil, Mexico & India Unemployment Rates by Year", x = "Date", y = "Percent", color = "Country") + 
  theme_classic()


```
	
In the image above, you will see the average annual unemployment rates for four countries. This measures the percent of people who cannot find a job in the group of those people either working or looking for work.

This is measured as 

$$ \text{Unemployment Rate} = \frac{\text{Unemployed}}{\text{Unemployed} + \text{Employed}} \times 100 $$

This material will accompany [@tw] which will be the primary book for the course. We will also be using [@core], and the other materials compiled by CORE including [@espp] and [@doing] which act as useful comparisons to the more traditional material presented in [@tw]. 

In recent months the COVID-19 crisis has gripped the world and put our global economy in a precarious position. As we can see from Figures \@ref(fig:jobs) and \@ref(fig:vmt) people who are newly jobless and travel in the U.S. have moved dramatically in opposite directions. You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r jobs, echo = FALSE, fig.cap='Jobless Claims Skyrocket in 2020', out.width='80%', fig.asp=.75, fig.align='center'}
jobs <- tq_get("ICSA", 
                 get = "economic.data",
                 from = Sys.Date() - lubridate::years(25))
jobs %>% ggplot(aes(date,price/1000)) + 
  geom_line() + 
  labs(x = "Date", y = "Thousands of Claims", title = "Initial Jobless Claims", color = "Series")
```

```{r vmt, echo = FALSE, fig.cap='Travel Collapses in 2020', out.width='80%', fig.asp=.75, fig.align='center'}
vmt <- tq_get("TRFVOLUSM227SFWA", 
                 get = "economic.data",
                 from = Sys.Date() - lubridate::years(25))
vmt %>% ggplot(aes(date,price)) + 
  geom_line() + 
  labs(x = "Date", y = "Millions of Miles", title = "Vehicle Miles Traveled", color = "Series")
```

Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```


