# Introduction {#intro}

```{r, include = FALSE}
library(readr)
library(quantmod)
library(zoo)
library(tidyquant)
library(tidyverse)
library(lubridate)
library(reshape2)
library(dplyr)
library(wbstats)
library(ggplot2)
library(usmap)
library(gtrendsR)
library(tmap)
library(spData)
library(pwt9)
library(googleVis)

```
	
```{r, echo = FALSE}
unems <- wb(country = c("BRA","IND","MEX","USA"),
               indicator = c("SL.UEM.TOTL.ZS"), mrv=30, gapfill=TRUE, POSIXct = TRUE)

ggplot(unems, aes(x = date_ct, y = value, color = iso2c)) + 
  geom_line(size = 1) +
  labs(title = "U.S., Brazil, Mexico & India Unemployment Rates by Year", x = "Date", y = "Percent", color = "Country") + 
  theme_classic()


```
	
In the image above, you will see the average annual unemployment rates for four countries. This measures the percent of people who cannot find a job in the group of those people either working or looking for work.

This is measured as 

$$ \text{Unemployment Rate} = \frac{\text{Unemployed}}{\text{Unemployed} + \text{Employed}} \times 100 $$

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
